Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support).

### 2017-02-10 csf-2017/paper.tex
added comments for macr section

The basic idea of our approach is quite simple: we stick to a single, rather
standard C stack and register-passed return pointers in much the same way as a
standard C ABI. However, to prevent various ways of misusing this basic scheme,
we put local capabilities to work and take a number of not-always-obvious safety
measures. In the next paragraphs, we will explain the issues to be considered in
all the relevant situations: when (1) starting our program, (2) Returning to the
adversary, (3) invoking the
adversary, (4) returning from the adversary, (5) invoking an adversary callback
and (6) having a callback invoked by the adversary.

@@

\textbf{Returning to the adversary} Perhaps the simplest scenario is the case of
returning to the adversary after they invoked our code. In this case, we will
have received a return pointer from him, and we just need to jump to it as
usual. The only thing we need to take care of is to properly clear the registers
before we do (erasing any data we do not want to pass as (part of) the return
value) and similarly, erase any part of the stack that we have used (i.e.
overwrite it with zeros). Note that we can use the stack as normal: we may push
values onto it and pop them back off (e.g. for register spilling etc.).

@@

Already important is the fact that the return pointer must be provided to the
adversary as a local capability. If not, the adversary would be able to store
away the return pointer in a global data structure, and he would be able to jump
to it in circumstances where this should not be possible: he may, for example,
store the return pointer, legally jump to it a first time, wait to be invoked
again and then jump to the old return pointer a second time, rather than the new
return pointer received for the new invocation. Similarly, he may store the
return pointer, invoke a function in our code, wait for us to invoke him again
and then jump to the stored, old return pointer rather than the new one he
received upon the second invocation.

@@

A final required safety measure is that before invocation of the adversary, we
need to clear (i.e. overwrite with zeros) the part of the stack that we are not
using. Otherwise, that part of the stack may still contain data and capabilities
from previously executed code, including local capabilities since they are
allowed on the stack. The adversary could take advantage of this by using it to
store either stack pointers or return pointers from previous invocations and
play the same tricks as above. Clearing the entire unused part of the stack is
the most problematic aspect of our discipline in terms of efficiency, since it
may consist of a large part of memory. We believe that capability machines like
CHERI should consider to provide special support for this requirement, by
implementing an optimized instruction for clearing (i.e. overwriting with zeros)
a large block of memory. We have no experience in processor design, but we
imagine this might be optimized as a vector instruction or by having the
processor keep track of a range of memory for which loads should return zero
irrespective of the actual contents of the memory.

@@

\paragraph{Having a callback invoked by the adversary}
The above leaves us with probably the hardest scenario: how to provide a
callback to the adversary. The basic idea is that we allocate a block of memory
using malloc that we fill with the capabilities and data that the callback needs
access to, as well as some prelude instructions that load the data into
registers and jumps to the right code. We can then provide the adversary with an
enter-capability covering the allocated block and pointing to the contained
prelude instructions. However, the question that remains in this setup is: where
do we get the stack pointer?

Our answer is that the adversary should provide it to us, much like we provide
him with a current stack pointer when we invoke his code.  However, it is
important that we do not just accept any capability as a stack pointer but check
that it is safe to use.  Specifically, we check that it is indeed a $\rwlx$
pointer and that its current address points just below its start address.
Without these checks, err...
\dominique{is this needed for security or just for convenience?}

@@

LB: maybe we can present the examples informally before the logical relation and then return to them after the logical relation (to present the formal properties) ?

>>>>

The basic idea of our approach is quite simple: we stick to a single, rather
standard C stack and register-passed return pointers in much the same way as a
standard C ABI\lau{This sentence does not compute}. However, to prevent various ways of misusing this basic scheme,
we put local capabilities to work and take a number of not-always-obvious safety
measures. In the next paragraphs, we will explain the issues to be considered in
all the relevant situations: when (1) starting our program, (2) Returning to the
adversary, (3) invoking the
adversary, (4) returning from the adversary, (5) invoking an adversary callback
and (6) having a callback invoked by the adversary.

@@

\textbf{Returning to the adversary} Perhaps the simplest scenario is the case of
returning to the adversary after they invoked our code. In this case, we will
have received a return pointer from them, and we just need to jump to it as
usual. The only thing we need to take care of is to properly clear the registers
before we do (erasing any data we do not want to pass as (part of) the return
value) and similarly, erase any part of the stack that we have used (i.e.
overwrite it with zeros). Note that we can use the stack as normal: we may push
values onto it and pop them back off (e.g. for register spilling etc.).

@@

Already important is the fact that the return pointer must be provided to the
adversary as a local capability. If not, the adversary would be able to store
away the return pointer in a global data structure, and he would be able to jump
to it in circumstances where this should not be possible: for example, he may
store the return pointer, legally jump to it a first time, wait to be invoked
again and then jump to the old return pointer a second time, rather than the new
return pointer received for the new invocation. Similarly, he may store the
return pointer, invoke a function in our code, wait for us to invoke him again
and then jump to the stored, old return pointer rather than the new one he
received upon the second invocation. \lau{If we need space, then maybe one of the examples is sufficient.}

@@

A final required safety measure is that before invocation of the adversary, we
need to clear (i.e. overwrite with zeros) the part of the stack that we are not
using. Otherwise, that part of the stack may still contain data and capabilities
from previously executed code, including local capabilities since they are
allowed on the stack. The adversary could take advantage of this by using it to
store either stack pointers or return pointers from previous invocations and
play the same tricks as above. Clearing the entire unused part of the stack is
the most problematic aspect of our discipline in terms of efficiency, since it
may consist of a large part of memory. We believe that capability machines like
CHERI should consider to provide special support for this requirement, by
implementing an optimized instruction for clearing
a large block of memory. We have no experience in processor design, but we
imagine this might be optimized as a vector instruction or by having the
processor keep track of a range of memory for which loads should return zero
irrespective of the actual contents of the memory.

@@

\textbf{Having a callback invoked by the adversary}
The above leaves us with probably the hardest scenario: how to provide a
callback to the adversary. The basic idea is that we allocate a block of memory
using malloc that we fill with the capabilities and data that the callback needs
access to, as well as some prelude instructions that load the data into
registers and jumps to the right code. We can then provide the adversary with an
enter-capability covering the allocated block and pointing to the contained
prelude instructions. However, the question that remains in this setup is: where
do we get the stack pointer?

Our answer is that the adversary should provide it to us, much like we provide
him with a current stack pointer when we invoke his code.  However, it is
important that we do not just accept any capability as a stack pointer but check
that it is safe to use.  Specifically, we check that it is indeed a $\rwlx$
pointer and that its current address points just below its start address.
Without these checks, err...
\dominique{is this needed for security or just for convenience?}
\lau{Say we are called with a stack capability that is global and $\rwx$ as well as an integer as a non-local return pointer (this is fair game as the adversary has to follow no rules). Now you push the l(a)unch codes to what you think is the stack and \texttt{scall} the callback. At this point \texttt{scall} does its thing, but because it never attempts to store anything local it will succeed and jump to the adversary! The adversary can now look at what you thought was the stack and eat your lunch or fire the missiles as they prefer. It is not sufficient to require the stack to be just local because there may be a global equivalent. We know this is not the case for the rwlx capability because we have a system wide assumption that all rwlx capabilities are local.
Finally the adjustment of the pointer in \texttt{prepstck} (or whatever I called it) is a convenience/ space efficiency thing purely.}

@@

\lars{maybe we can present the examples informally before the logical relation and then return to them after the logical relation (to present the formal properties) ?}
